---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggthemes)
library(tidycensus)
```



## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

Done


```{r}

od_df <- read_csv("data/prince_georges_2022_overdoses.csv")

od_df <- od_df %>%
  mutate(call_day = day(datetime),
         call_week = week(datetime),
         call_month = month(datetime, label = TRUE))

```



## Questions 

**Q1.** Which month saw the greatest percentage of total calls?

Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.**

Dec had the highest percentage of total calls.

This is very different than what we saw in Baltimore County in the pre-lab, which has the highest calls in Mar-May and Aug instead


```{r}

monthly_od_df <- od_df %>%
  group_by(call_month) %>%
  summarise(monthly_calls = n()) %>%
  mutate(percent_of_yearly_calls = monthly_calls / sum(monthly_calls) * 100) %>%
  arrange(desc(percent_of_yearly_calls))
  

monthly_od_df

```

```{r}


monthly_od_df %>%
  ggplot() +
  geom_bar(aes(x=call_month, weight=percent_of_yearly_calls), fill ="red", color="black") +
  theme_economist() +
  labs(
    title="PG County Overdoses Highest in July and Nov-Dec",
    x = "Month",
    y = "Percent of Annual 911 Overdose Calls",
    caption = "source: PG County EMS"
    
  )


```


**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Done

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

Done

**A2.** 

There are pretty regular peaks and valleys, but the highest totals are in July and the Holiday Season. I'd be curious to dive into other granularities of date-based representations of the data to see if the fluctuations align with something like bi-weekly paychecks.

Week 53 is a huge outlier. That's because week 53 only contains one day (Dec 31) instead of 7 days like all he other weeks, so naturally there are fewer total calls. I would either exclude week 53 from the visualization, or do something to weight the values so that we're not comparing total calls in a one day week against seven day weeks.


```{r}

weekly_od_df <- od_df %>%
  group_by(call_week) %>%
  summarise(weekly_calls = n())

weekly_od_df

```

I asked chatGPT how to update my code below to make the x axis have ticks at intervals of 5 instead of 20. My prompt was: "how do I update the following code to make the x label ticks 5 at a time? [code]"

```{r}

weekly_od_df %>%
  ggplot() +
  geom_line(aes(x=call_week, y=weekly_calls), color="purple") +
  theme_economist() +
  labs(
    title="PG County Overdoses Highest in July and Nov-Dec",
    x = "Week of Year",
    y = "Calls Per Week",
    caption = "source: PG County EMS"
    ) +
  scale_x_continuous(breaks = seq(min(weekly_od_df$call_week), max(weekly_od_df$call_week), by = 5))

```


```{r}
acs5_variables <- load_variables(2021, "acs5", cache = TRUE)
acs5_variables
```



**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.

```{r}

zip_od_df <- od_df %>%
  group_by(zipcode) %>%
  summarise(calls_per_zip = n())

zip_od_df

```


2. Get data from the Census Bureau showing median household income for Maryland zip codes.

```{r}

hhincome_df <- get_acs(geography = "zcta",
              variables = c(medage = "B19013_001"),
              state = "MD",
              year = 2019) %>%
      arrange(estimate)

hhincome_df

```


3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.

```{r}

zip_od_hhincome <- zip_od_df %>%
  inner_join(hhincome_df, join_by("zipcode" == "GEOID"))

zip_od_hhincome

```


4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.


```{r}

zip_scatter <- zip_od_hhincome %>%
  ggplot() +
  geom_point(aes(x=estimate, y=calls_per_zip), color="blue") +
  xlab("Median Household Income Per Zip") +
  ylab("Number of Overdose Calls Per Zip") +
  labs(
    title = "Zip Codes with higher HH Income have smaller, less varied overdose totals",
    caption = "Sources: 2019 ACS, PG County EMS \n Each point represents a Zip Code in PG County") +
  geom_text(aes( x=estimate+4500, y=calls_per_zip, label = zipcode), check_overlap = TRUE)


zip_scatter

```



5. Give it an appropriate title, source, and x and y axis titles.

Done

6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.

Done. Note - I needed chatGPT's help to tell me that I needed to redefine x and y values for geom_text rather than it piggybacking off the x and y in geom_point(). I sent it my error message and code and asked it to explain the error message.


7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  

There's a horizontal funnel-like trend in the data -- there's HETEROSCEDASTICITY! (Disclaimer: I will forever need to google how to spell that). Zip Codes with higher HH Income have smaller, less varied overdose totals than less-wealthy zips.

This makes it a little more difficult to tease out the relationship between HH Income and OD totals. That's interesting and fun from a statistical standpoint and means we should play a little more with the data -- for example, would we see homoscedasticity instead if we used OD calls per capita instead of raw totals?

But we should never write and article about heteroscedasticity, ever.

Instead, I would write an article diving into the question of WHY there's so much variation in the total # of overdoses in less-wealthy zip codes. Use case studies to illustrate the issue -- interview people and do research specifically on 20608 which has a very low total # of overdoses, versus 20743 which has a MUCH higher # of ODs despite a pretty similar median hh income. If the difference isn't due to population size differences, what can we learn from the wild variation re: what factors do/don't affect OD numbers?
