---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

# Load janitor
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

Load the data

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564)

# Load in the Maryland grants and loan data

md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

```

Clean the column names using janitor

```{r}
library(janitor)

# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()

# display the cleaned dataset
cleaned_md_grants_loans
```

```{r}
cleaned_payments = payments %>%
  clean_names()

cleaned_payments
```


## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1:

I found 87 records

```{r}
# I'm going to interpret this question as asking me to include payments to PG County's Board of Education, too


# Let's start with just "PRINCE GEORGE"

prince_george_filter = cleaned_payments %>%
  filter(str_detect(vendor_name, "PRINCE GEORGE"))

prince_george_filter

# Possible values to match on: PRINCE GEORGES COUNTY PUBLIC SCHOOL, PRINCE GEORGES COUNTY PUBLIC SCHOOLS, PRINCE GEORGES COUNTY SCHOOL, PRINCE GEORGES PUBLIC SCHOOLS,


# Also: BOARD OF EDUCATION OF PRINCE GEORGE, BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD, BOARD OF EDUCATION OF PRINCE GEORGE'S COUNTY



# Let's check "PG" too

pg_filter = cleaned_payments %>%
  filter(str_detect(vendor_name, "PG"))

pg_filter

# Possible values to match on: PG COUNTY PUBLIC SCHOOL, BD OF EDUCATION OF PG CO

```


```{r}
# Construct the data frame "pg_schools" to only include the vendor names above


pg_schools = cleaned_payments %>%
  filter(str_detect(vendor_name, "PRINCE GEORGE") | str_detect(vendor_name, "PG")) %>%
  filter(
    str_detect(vendor_name, "SCHOOL"
    ) |
    str_detect(vendor_name, "BOARD OF EDUCATION"
    ) |
    str_detect(vendor_name, "BD OF EDUCATION OF PG CO"
    ))

pg_schools


```



Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:

FY 2022 had the largest total amount. Something that stands out to me: as a very general statement, the amount spent generally tends to increase year after year, with an important exception: 2010-2014 are the lowest years on record, and are lower than 2008-2009 spending levels. I wonder what might've caused this -- perhaps it's tied to cutbacks during the post '08-09 recession period?

```{r}
pg_schools %>%
  group_by(fiscal_year) %>%
  summarise(total_year_amount = sum(amount)) %>%
  arrange(desc(total_year_amount))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.


- Some 5 digit zip codes for seemingly the same vendor are different. Both 21201 and 21204 are listed for the UMD Medical Center (unless my assumption is wrong about UMD Medical Center and UMD St/St. Joseph Medical Center being the same thing)

Maybe it would be more useful to only use the first three digits of zip codes -- create a new column with those Zip-3's and group on that instead of vendor_zip


- Vendor names aren't standardized. UNIVERSITY OF MARYLAND ST JOSEPH MEDICAL CENTER is certainly the same thing as UNIVERSITY OF MARYLAND ST. JOSEPH MEDICAL CENTER. I'm assuming again that those are also the same thing as UNIVERSITY OF MARYLAND MEDICAL CENTER in 21201.

Solution: Removing punctuation from the values in vendor_name would solve the issue of "ST" vs "ST." To match vendor names with ST/ST. JOSEPH with those that are just "UNIVERSITY OF MARYLAND MEDICAL CENTER," and if we're ONLY working to clean those three specific vendor names, you could pretty efficiently use a case when to standardize the names. That would become much more tedious of a solution though if we're trying to use case when to standardize ALL names in the original df. In that scenario, maybe fuzzy matching?...


```{r}

umd_payments <- cleaned_payments %>%
  filter(str_detect(agency_name, "CAPITAL PROJECTS")) %>%
  filter(str_detect(vendor_name, "UNIVERSITY OF MARYLAND")) %>%
  group_by(vendor_name, vendor_zip) %>%
  summarise(total_amount_spent = sum(amount))

umd_payments

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4:

Stem cell research was at its highest in 2009, after which funding was roughly halved and then stays between high 7mil to low 9 mil of funding from 2009-2013. Funding drops considerably in 2014, and seemingly NO grants were awarded in 2016-17. Funding returns to 2014ish levels in 2018 and stays in the 7,000,000s through 2022, with the exception of a huge single-year drop in 2021

```{r}
# values in the description column aren't standardized as upper/lower case. Need to fix this first so str_detect doesn't miss any matches. Also doing that for grantee names so group_by in Q5 doesn't mismatch upper and lower case versions of the same grantee

cleaned_md_grants_loans$description = str_to_upper(cleaned_md_grants_loans$description)
cleaned_md_grants_loans$grantee = str_to_upper(cleaned_md_grants_loans$grantee)

# Now let's use str_detect to filter the df to rows with "stem cell" in the description

stem_cell_grants = cleaned_md_grants_loans %>%
  filter(str_detect(description, "STEM CELL RESEARCH GRANT")) %>%
  group_by(fiscal_year) %>%
  summarise(amount_spent = sum(amount))

stem_cell_grants


```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: 

It looks like UMD College Park records are associated with either 20740 or 20742 zip codes, so I'm using both of those.

Similar to Q&A 3, some of the mismatches could be solved by removing punctuation and/or whitespaces, and using str_to_upper on the grantee names (I've gone back and done that last part during Q3 retroactively.)

After that, either fuzzy matching or case when plus str detect to standardize the vendor names in this specific list?

The most interesting thing to me is that there are almost no individuals who were grantees, the only one I see is ELIZABETH GALLAURESI. I would be curious to look into whether this is because very few individuals apply for grants, or if it's because very few are awarded them.


```{r}

cleaned_md_grants_loans %>%
  filter(zip_code == "20740" | zip_code == "20742") %>%
  group_by(grantee) %>%
  summarise(amount_awarded = sum(amount)) %>%
  arrange(grantee)


```
```{r}

cleaned_md_grants_loans %>%
  filter(zip_code == "20740" | zip_code == "20742") %>%
  group_by(grantee) %>%
  summarise(amount_awarded = sum(amount)) %>%
  arrange(desc(amount_awarded))

```

