---
title: "lab_10"
author: "Derek Willis"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)
library(janitor)
library(tigris)
library(sf)
library(tidycensus)


```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history last November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.**

Prince George's County and Baltimore City have the most interesting results -- the story here is that only in those two locations, Brown earned a higher percent of the vote in the ag race than Moore earned in the gov race. It's worth noting though that Moore still earned more raw votes than Brown did in those counties.

```{r}

md_election <- read_csv("data/md_statewide_county.csv") %>%
  clean_names()

md_election

```

```{r}
md_election <- md_election %>%
  mutate(total_gov_votes = cox + moore + lashar + wallace + harding + sefcik + gov_writein) %>%
  mutate(total_ag_votes = peroutka + brown + ag_writein) %>%
  mutate(moore_pct = moore / total_gov_votes * 100) %>%
  mutate(brown_pct = brown / total_ag_votes * 100) %>%
  mutate(moore_diff_v_brown = moore_pct - brown_pct) %>%
  mutate(geoid = as.character(geoid))

md_election
```

```{r}
counties <- counties()
md_counties <- counties |>
  filter(STATEFP == "24")

md_counties
```

```{r}
md_election_w_geo <- md_election %>%
  left_join(md_counties, join_by("geoid" == "GEOID")) %>%
  clean_names()

md_election_w_geo
```

```{r}
county_centroids <- st_centroid(md_counties)
county_centroids_df <- as.data.frame(st_coordinates(county_centroids))
county_centroids_df$NAME <- county_centroids$NAME
```


```{r}

ggplot() +
  geom_sf(data=md_election_w_geo, aes(fill=moore_diff_v_brown, geometry=geometry)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE, color = "orange") +
  theme_minimal() +
  scale_fill_viridis_b(option="magma") +
  xlab("Latitude") +    # Set x-axis label
  ylab("Longitude")    # Set y-axis label

# I asked chatGPT how to update the color of the text of the labels. "orange" still isn't great here but it does allow the labels to be visible for the two counties I found the most interesting (PG County and Baltimore City), and I think most of the other label colors I tried were even worse. I'm sticking with magma because I don't think most of the color scale options intuitively imply which colors are high vs low. I'm also not using the default blue-gradient color scale you get if you don't use scale_fill_viridis_b() because I don't think it effectively/sufficiently portrays the significant difference between the highest and lowest counties here.

# If I had time to learn how to do this, I would keep all the labels black EXCEPT for those two counties, which I would make a different color

# I also tried out a second version below where the color scale is inverted (also with chatGPT's help). Some other counties have their labels hidden as a result but PG County and Baltimore City's are more visible there. I don't like it as much, though.

# I ALSO (last time) asked ChatGPT how to update the axis labels instead of just showing X and Y
```

```{r}

ggplot() +
  geom_sf(data=md_election_w_geo, aes(fill=moore_diff_v_brown, geometry=geometry)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma", direction = -1) +
  xlab("Latitude") +    # Set x-axis label
  ylab("Longitude")    # Set y-axis label

```


**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.**

Talbot had a MUCH higher rate of outages than any other county -- while Washington and Frederick had relatively few outages. If I had access to more months' data, I would want to see if those trends continue or if Talbot's huge rate might have been a one-off occurrence due to some event that negatively impacted the county (eg a bad storm or tornado)? If those big discrepancies are consistent though, I would want to dig further into potential reasons -- does Talbot county have fewer resources than other counties? Are there environmental conditions that make power outages more likely there?


```{r}

outage_df <- read_csv("data/county_outages_march2023.csv")

outage_df

```


```{r}
# update the county names in md_counties so we can join on that

md_counties
```

```{r}

# update Baltimore city's name to allow for joining on name

md_counties$NAME[3] = "Baltimore City"

```


```{r}
md_counties
```



```{r}

outage_df <- outage_df %>%
  mutate(outage_per_10k = outage_customers / total_customers * 10000)

outage_df

```

```{r}

outages_w_geo <- md_counties %>%
  full_join(outage_df, join_by("NAME" == "county"))

outages_w_geo

```

```{r}
ggplot() +
  geom_sf(data=outages_w_geo, aes(fill=outage_per_10k, geometry=geometry)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE, color = "orange") +
  scale_fill_viridis_b(option="A", trans = "log") +
  xlab("Latitude") +    # Set x-axis label
  ylab("Longitude")    # Set y-axis label
```


**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.**

Caroline county has way more than evs and hybrids than any of the others! I would be curious to see if there were any county-level incentives introduced there for ev ownership, or more charging stations installed in the county that makes ev use more convenient than before, etc. Meanwhile, Dorchester has the lowest per capita change -- has there been almost no change in the number of charging stations? Or is there something else about Dorchester that makes ev/hybrid ownership less practical and less likely to change? Eg, is it a rural or agricultural county where trucks are still needed more than hybrids or evs for work/lifestyle reasons?

The top and lowest counties are different though when you analyze it by per capita ownership rather than by % change. Montgomery and Howard have the highest, while Somerset has the lowest.

```{r}

ev_df <- read_csv("data/ev_2020_2023.csv") %>%
  mutate(percent_change = (march_2023 - july_2020) / july_2020 * 100)

ev_df

```


```{r}
#join w geo data

ev_w_geo <- md_counties %>%
  left_join(ev_df, join_by("NAME" == "county"))

ev_w_geo


```

```{r}

ggplot() +
  geom_sf(data=ev_w_geo, aes(fill=percent_change, geometry=geometry)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE, color = "orange") +
  scale_fill_viridis_b(option="A", trans = "log") +
  xlab("Latitude") +    # Set x-axis label
  ylab("Longitude")    # Set y-axis label

```



```{r}
#get the pop data for MD counties from ACS

md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD")

md_county_population
```


```{r}
# join with ev_w_geo

ev_w_pop_geo <- ev_w_geo %>%
  left_join(md_county_population, by = join_by(GEOID))

ev_w_pop_geo


```


```{r}
ev_w_pop_geo <- ev_w_pop_geo %>%
  mutate(ownership_per_10k = march_2023 / estimate * 10000)

ev_w_pop_geo
```


```{r}

ggplot() +
  geom_sf(data=ev_w_pop_geo, aes(fill=ownership_per_10k, geometry=geometry)) +
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, color = "orange") +
  scale_fill_viridis_b(option="A", trans = "log") +
  xlab("Latitude") +    # Set x-axis label
  ylab("Longitude")    # Set y-axis label

```

